---
title: 'Final Presentation: Group 1'
author: "Mia Josephy-Zack, Alex Medeiros, Mila Lewis-Peinado"
output: ioslides_presentation
date: "2025-05-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(ggplot2)
library(tidyr)

squirrels <- read.csv("squirrel-data.csv")
parks <- read.csv("park-data.csv")
tree <- read.csv("NYC-parkdata.csv")
data <- left_join(squirrels, parks, by = c("Park.ID", "Park.Name", "Area.Name", "Area.ID"))
full_data <- left_join(data, tree, by = c("Park.ID", "Park.Name"))

# cleaning data: converting char to num
full_data$Total.Acres <- as.numeric(full_data$Total.Acres)
full_data$Number.of.Mapped.Trees <- as.numeric(full_data$Number.of.Mapped.Trees)
full_data$Acres.of.Natural.Areas <- as.numeric(full_data$Acres.of.Natural.Areas)

# cleaning data: converting char to num
full_data$Total.Acres <- as.numeric(full_data$Total.Acres)
full_data$Number.of.Mapped.Trees <- as.numeric(full_data$Number.of.Mapped.Trees)
full_data$Acres.of.Natural.Areas <- as.numeric(full_data$Acres.of.Natural.Areas)

# cleaning data cont.: cleaning up the comments

# park condition variable
full_data$Park.Conditions[full_data$Park.Conditions == "Calm, 20\x8a\x97\x9630 ppl on each side"] <- "Calm"
full_data$Park.Conditions <- ifelse(
  grepl("calm", full_data$Park.Conditions, ignore.case = TRUE),
  "Calm",
  ifelse(
    grepl("busy", full_data$Park.Conditions, ignore.case = TRUE),
    "Busy",
    NA
  )
)

# litter variable
full_data$Litter <- ifelse(
  grepl("some", full_data$Litter, ignore.case = TRUE),
  "Some",
  ifelse(
    grepl("abundant", full_data$Litter, ignore.case = TRUE),
    "Abundant",
    "None"
  )
)

# human interaction variable
# I coded the default to be indifferent, bc the ones that didn't mention
# any keywords (runsfrom/approaches) sounded like indifference

full_data$Interactions.with.Humans <- ifelse(
  grepl("runs from", full_data$Interactions.with.Humans, ignore.case = TRUE),
  "Runs from",
  ifelse(
    grepl("approaches", full_data$Interactions.with.Humans, ignore.case = TRUE),
    "Approaches",
    "Indifferent"
  )
)

# rate variable: squirrels per acres

# need to account for NA value (park with no total acres)

full_data <- full_data %>%
  mutate(Squirrel.rate = Number.of.Squirrels / Total.Acres)

# Rename latitude and longitude variables
full_data <- full_data %>%
  rename(
    lat = "Squirrel.Latitude..DD.DDDDDD.",
    long = "Squirrel.Longitude...DD.DDDDDD."
  )

full_data$Squirrel.rate <- as.numeric(full_data$Squirrel.rate)

# cleaning activities variable
core_activities <- c("Foraging", "Eating", "Running", "Climbing", "Chasing", "Digging", "Sitting", "Sleeping")

extract_activities_grepl <- function(activity_str) {
  # na if missing
  if (is.na(activity_str) || activity_str == "") return(NA)
  
  #checking for word
  found <- core_activities[sapply(core_activities, function(act) grepl(act, activity_str, ignore.case = TRUE))]
  
  if (length(found) == 0) return(NA)
  paste(found, collapse = ", ")
}

full_data$Cleaned.Activities <- sapply(full_data$Activities, extract_activities_grepl)

# cleaning animal sightings variable
animal_list <- c("dogs", "cat", "pigeons", "cardinals", "hawks", "rat", 
                 "robins", "sparrows", "goose", "downy woodpecker", 
                 "song birds", "fish", "humans", "geese")

pattern <- paste(animal_list, collapse = "|")

clean_animal_sighting_grepl <- function(sighting) {
  if (is.na(sighting) || sighting == "") return(NA)
  sighting <- gsub("\\([^\\)]*\\)", "", sighting)
  sighting <- tolower(sighting)
  matches <- unique(unlist(regmatches(sighting, gregexpr(pattern, sighting))))
  if (length(matches) == 0) return(NA)
  paste(matches, collapse = ", ")
}

full_data$Cleaned.Animal.Sightings <- sapply(full_data$Other.Animal.Sightings, clean_animal_sighting_grepl)

#make rate group variable
full_data <- full_data %>%
  mutate(Rate.Group = case_when(
    Squirrel.rate < 1.206 ~ "Low",
    Squirrel.rate < 2.045 ~ "Medium",
    Squirrel.rate < 5.231 ~ "High",
    TRUE ~ "Very High"
  ))
```

## Research Question

- How do squirrel-human interactions differ across various NYC parks?
  - Interested in environmental and atmosphere/ambiance factors
  - Squirrels are important part of the NYC ecosystem & health of urban green spaces
- Our hypothesis: Busier parks will have braver squirrels!
  - Adaptation from exposure 

## Our Dataset

- Dataset source: The Squirrel Census sample from March 1st, 2020
- 3 Datasets: Park dataset, Squirrel dataset, & Park Info dataset (NYCgovparks.org, made by us)
  - Left-joined by common variables (area name, area ID, park name, and park ID)
- Cleaned data mainly using grepl() function
  - for some variables, people were allowed to write additional comments

## Our Dataset (cont.)

- Park Dataset:
  - Park conditions (boolean: calm or busy)
  - Area name: area where the park is located (upper manhattan, central manhattan, lower manhattan, or brooklyn)
  - Area ID: A, B, C, or D assigned to an area name
  - Park name: name of the park
  - Park ID: 1-24 assigned to park names
  - Number of squirrels: squirrel count at specific park
  - Litter (some, none, abundant)

## Our Dataset (cont.)

- Squirrel Dataset:
  - Area name, Area ID, Park name, Park ID
  - Primary fur color: gray, black, or cinnamon
  - Squirrel longitude: longitude coordinate for observed squirrel
  - Squirrel latitude: latitude coordinate for observed squirrel
  - Interactions with Humans: approaches, indifferent, or runs from
  - Activities: value is either “running”, “chasing”, “climbing”, or “foraging” 
  - Location: location when first sighted (value is either “ground plane”, “above ground”, or specific location)

## Our Dataset (cont.)

- Park Info Dataset:
  - Park name
  - Park ID
  - Total acres
  - Acres of natural areas
  - Number of mapped trees
- Helped us group parks
- We manually created the following variables:
  - Squirrel Rate: Number of Squirrels / Total Acres
  - Rate Group: Quantiles of Squirrel Rate
  - Park Size: Quantiles of Total Acres

## Atmosphere/ambiance: Stacked Barplot

```{r, echo=FALSE, message=FALSE}
interaction_counts <- full_data %>%
  group_by(Park.Conditions, Rate.Group, Interactions.with.Humans) %>%
  summarise(count = n()) %>%
  filter(!is.na(Park.Conditions))

ggplot(interaction_counts, aes(x = Park.Conditions, y = count, fill = Interactions.with.Humans)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~Rate.Group, scales = "free_x") +
  labs(title = "Squirrel-Human Interactions by Park Condition (Grouped by Squirrel Rate)",
       x = "Park", y = "Count") + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Atmosphere/ambiance: Heatmap

```{r , echo=FALSE, message=FALSE}
animals_long <- full_data %>%
  separate_rows(Cleaned.Animal.Sightings, sep = ",\\s*") %>%
  group_by(Cleaned.Animal.Sightings, Interactions.with.Humans) %>%
  summarise(count = n()) %>%
  filter(!is.na(Cleaned.Animal.Sightings) & Cleaned.Animal.Sightings != "")

ggplot(animals_long, aes(x = Cleaned.Animal.Sightings, y = Interactions.with.Humans, fill = count)) +
  geom_tile() +
  labs(title = "Heatmap: Interactions vs. Other Animal Sightings",
       x = "Other Animal Sightings", y = "Squirrel-Human Interaction") +
  scale_fill_gradient(low = "white", high = "steelblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Environment: Heatmap

```{r , echo=FALSE, message=FALSE}
full_data$Total.Acres <- as.numeric(full_data$Total.Acres)

full_data <- full_data %>% 
  filter(!is.na(Total.Acres)) %>%
  mutate(Park.Size = cut(Total.Acres, breaks = quantile(Total.Acres, probs = c(0, 1/3, 2/3, 1)), labels = c("Small", "Medium", "Large")))

environment_heatmap_data <- full_data %>%
  group_by(Litter, Interactions.with.Humans, Park.Size) %>%
  summarise(Tot.count = n()) %>%
  filter(!is.na(Park.Size))

ggplot(environment_heatmap_data, aes(x = Interactions.with.Humans, y = Litter, fill = Tot.count)) + geom_tile() + labs(x = "Squirrel-Human Interaction", y = "Litter Presence", fill = "Number of Observations", title = "Environmental Factors: Heatmap of Park Size and Litter Presence") + theme_minimal() + scale_fill_gradient(low = "brown", high = "green3") + geom_text(aes(label = paste(Tot.count)), size = 3) + facet_wrap(~Park.Size) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Environment: Heatmap (cont.)

```{r , echo = FALSE, message = FALSE}
# sorts them by park rate

environment_heatmap_data <- full_data %>%
  group_by(Litter, Interactions.with.Humans, Rate.Group) %>%
  summarise(Tot.count = n()) %>%
  filter(!is.na(Rate.Group))

ggplot(environment_heatmap_data, aes(x = Interactions.with.Humans, y = Litter, fill = Tot.count)) + geom_tile() + labs(x = "Squirrel-Human Interaction", y = "Litter Presence", fill = "Number of Observations", title = "Environmental Factors: Heatmap by Rate Group") + theme_minimal() + scale_fill_gradient(low = "brown", high = "green3") + geom_text(aes(label = paste(Tot.count)), size = 3) + facet_wrap(~Rate.Group)
```

## Squirrel Map: Shiny App

- Parks in Lower Manhattan tend to be busier than parks in Upper Manhattan.
  - We also observe that squirrels in Lower Manhattan are more likely to approach humans than those in Upper Manhattan.
- Many had missing data for longitude and latitude, so we estimated using different methods.
- [https://admedeiros.shinyapps.io/Final_Project/](https://admedeiros.shinyapps.io/Final_Project/)



